LME-MPM applied to quasi-brittle fracture.

### 2020-05-26 2020_EFM_MPM_Eigensoftening.tex

\newglossaryentry{nu}
{
  name=$\nu$,
  description={Poisson ratio}
}

@@

\subsection{The \acrshort{mpm} time integration : A \acrlong{npc}  scheme}
\label{sec:2.1}
The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian
method where moving material points, henceforth in this research,
particles \MMP{($\square_p$)} will carry on all the physical information of the local state ($\sigma_p,
\varepsilon_p$) and a set of fixed background nodes \MMP{($\square_I$)} will
be introduced to compute the balance of momentum equation. Since the
\acrshort{mpm} possesses the advantages of both Lagrangian and
Eulerian descriptions, no element distortion takes place in the
\acrshort{mpm}. Therefore, it is an appropriate and efficient method
to solve problems with moving discontinuities such as fracture evolution.\\
Without loosing generality, the \acrshort{mpm} algorithm
can be described with three main steps: (i) a variational recovery
process, where particle data is projected to the grid nodes, (ii) an
Eulerian step, where balance of momentum equation is expressed as a
nodal equilibrium equation in a FEM-like procedure, and finally
(iii) a Lagrangian advection of the particles. In the present research,
an explicit predictor-corrector time integration scheme is adopted. The
purposes of this choice is motivated due to its proved robustness and
stability for dynamic computations. \MMP{In the first stage, the nodal
velocity predictor $ \vec{\tilde{v}}_I^{k+1}$ is computed thorough
nodal velocities $\vec{v}_I$ and accelerations $\vec{a}_I$ evaluated
in the $k$ time step},
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{\tens{m}_I^{k+1}}\ .
\end{equation}
\MMP{Where \gls{gamma-npc} is an user-defined parameter (typically adopted as
0.5), $\tens{m}_I$ represents the lumped mass matrix, and $N_{Ip}^{k}$ is
the nodal contribution of the shape function evaluated in the particle
coordinates $\vec{x}_p^k$.} This way of computing the nodal predictor is numerically stable
as well as minimises the computational effort. Once nodal velocities are
obtained, the essential boundary conditions are imposed. After that, an
Eulerian phase is computed in the set of nodes in a
\acrshort{fem}-like way, where nodal forces $\vec{f}_{I}^{k+1}$ are
computed through the equilibrium equation. Next the nodal velocities
are corrected in a \textit{corrector} stage:
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\
  \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}\ .
\end{equation}
Finally, in order to update the particles, those are advected in the Lagrangian stage as
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} =
          \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ .
\end{align}
The complete pseudo-algorithm it is summarised in \ref{sec:expl-pred-corr}.

@@

(0,\infty)$.\MMP{Notice that $\beta$ has units of $[L]^{-2}$,

@@

$\widehat{\gamma}=\beta h^2$}, where $h$ is defined as a

@@

function is that in practice the value of $N_I$ decay exponentially following $\exp(-\beta \vec{r}
)$. In this sense
a good practice is to truncate it under a tolerance, 10$^{-6}$ would ensure a reasonable

@@

\subsection{Fracture modelling approach}
\label{sec:2.3}
Within the context of \acrshort{mpm} formulation, fracture can be
modelled by failing particles according to a suitable failure
criterion. When material points are failed, they are assumed to have
null stress tensor. To reproduce this behaviour in the present
research, the eigensoftening algorithm is introduced in the 
\acrshort{mpm} framework as an alternative approach to the decohesion
model \cite{Zhenmao_2005,Schreyer_2002}. The eigensoftening concept
was originally developed by Navas {\it et al.}
(2017)\cite{Navas_2017_ES} as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. A
comparison between both in \cite{Navas_2017_ES} shows that the
eigenerosion algorithm significantly overestimates the tensile  
stress and the strain peaks, while it captures the forces and crack
patterns accurately. On the other hand eigensoftening algorithm agree
very well with experimental results in all the aspects. Furthermore,
this algorithm has proof its accuracy for complex fracture
patters such the present in fiber reinforces concrete (FRC),
\cite{Navas_2018_ES}.\\

The key idea behind the eigenerosion algorithm is the computation of the
energy-release rate attendant to the failure of material point $p$,
\begin{align}
  \label{eq:energy-release-EE}
&G_p^{k+1} = \frac{C_{\epsilon}}{m_p^{k+1}}  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q W_q^{k+1}\\
&m_p^{k+1} =  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q  
\end{align}
where $B_{\epsilon}(x_p^{k+1})$ is a $^n$dimensional sphere of radius $\epsilon$
centered at $x_p^{k+1}$. The particles which lies under this sphere are
known as the $\epsilon$-neighborhood of the material point, see
\cite{Pandolfi_2012}, this concept is conveniently is sketched in
Figure \ref{fig:Failed-particles}.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure-Particle-failed}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Failed-particles}
\end{figure}
%%%%%%%%%%%
Other parameters are, the mass of the neighborhood $m_p^{k+1}$, the
current free-energy density per unit mass  $W_q^{k+1}$ and finally a
normalizing constant $C_{\epsilon}$. 
The failure criterion consist in to consider the material point fails
when $G_p^{k+1}$ surpasses a critical energy release rate that
measures the material-specific energy, $G_F$. The convergence of this
approach has been analyzed by Schmidt {\it et al.}
(2009)\cite{Schmidt_2009}, who proofed that it converges to the Griffith
fracture when discretization size tends to zero. It is necessary to
point out that when a material point overpass the critical energy, its
contribution to the internal forces vector is set to zero, but its
contribution to the mass matrix is preserved.\\

@@

in to account the gradual failure in quasi-brittle materials. The idea

@@

cohesive crack appears when the maximum tensile strength, $f_t$ is
reached. Once the opening displacement $w$ reach the value of the
critical crack displacement $w_c$, a stress-free crack is
attained. For intermediate values, $w_n$, a damage value between zero

@@

  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} \delta \tens{\epsilon}_q,

@@

material point in the $\epsilon$-neighborhood. In this point in
introduced an effective strain $\varepsilon_q$, such the variation of
the local strain energy can be obtained as $\delta W_q = \sigma_{q,1}
\delta\varepsilon_q$. Now with the assumption that the effective

@@

Consequently it is possible to define an equivalent critical stress at the
material point $\vec{x}_p^{k+1}$as

@@

where $m_p$ is the total mass of the $\epsilon$-neighborhood
  \label{eq:averaged-mass}
  m_p = \sum_{x_q^{k+1} \in B_{\epsilon}(x_p^{k+1})} m_q.

@@

tensile strength, $f_t$. This consists in a reduction of the internal

@@

variable and the volume for each material point $p$. $\chi_p$ takes

@@

material points). For the case of a linear softening such the sketched in the Figure \ref{fig:Damage-ft-wc}, $\chi_p$ is computed as,

@@

In analogy to the band crack model presented by

@@

\cite{Navas_2017_ES} introduced a band width parameter $h_{\epsilon}$.
Concerning this parameter, a typical value of it is between two and
four times the maximum size of the aggregates, in the case of

@@

$\varepsilon_1(\vec{x}_p^{0})$, and the current strain, $\varepsilon_1(\vec{x}_p^{k+1})$, for material point $p$. Also,

@@

The function of $\chi$ presented in \eqref{eq:damage-variable-chi-III}

@@

\MMP{
\subsection{$\epsilon$-neighborhood reconstruction : A node-linked method}
\label{sec:epsil-neighb-reconst}
Since this operation could be extremely demanding, optimal
numerical implementation should be employed. In opposite with the Cell-linked
method proposed by Allen \& Tildesley \cite{Allen_et_al_1989} where
the definition of a numerical tolerance is required for auxiliar
algorithms like Crossing Number \cite{Shimrat_1962} or Correct
Even-Odd \cite{Galetzka_et_al_2017}. To avoid these error prone
techniques, the method here proposed exploits the meshfree benefits of
the \acrshort{lme} approximants to introduce a node-linked method. Due
to the global definition of this interpolation technique, the classical
interpretation of the \acrshort{fem} element typically adopted in
\acrshort{mpm} simulations is not required anymore, instead a linked
list of tributary nodes is reconstructed with the closest node as
starting point at each time step. Therefore, each node is assigned an
unique linked-list of those particles which are closest to it. This
approach allows to reconstruct easily a search list with the support
of the initial connectivity of the mesh and offers a suitable basis
for an easy implementation of other interpolation techniques with
minimum coding effort. A detailed explanation of the proposed
algorithm can be found in \ref{sec:node-linked-method-1}.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cases of study and discussion}
\label{sec:3}

>>>>

\newglossaryentry{nu}
{
  name=$\nu$,
  description={Poisson ratio}
}
\begin{itemize}
\item 
\end{itemize}

@@

\subsection{The \acrshort{mpm} time integration : A \acrlong{npc}  scheme}
\label{sec:2.1}
The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian
method where moving material points, henceforth in this research,
particles ($\square_p$) will carry on all the physical information of the local state ($\sigma_p,
\varepsilon_p$) and a set of fixed background nodes ($\square_I$) will
be introduced to compute the balance of momentum equation. Since the
\acrshort{mpm} possesses the advantages of both Lagrangian and
Eulerian descriptions, no element distortion takes place in the
\acrshort{mpm}. Therefore, it is an appropriate and efficient method
to solve problems with moving discontinuities such as fracture evolution.\\
Without loosing generality, the \acrshort{mpm} algorithm
can be described with three main steps: (i) a variational recovery
process, where particle data is projected to the grid nodes, (ii) an
Eulerian step, where balance of momentum equation is expressed as a
nodal equilibrium equation in a FEM-like procedure, and finally
(iii) a Lagrangian advection of the particles. In the present research,
an explicit predictor-corrector time integration scheme is adopted. The
purpose of this choice is motivated due to its proved robustness and
stability for dynamic computations. In the first stage, the nodal
velocity predictor $ \vec{\tilde{v}}_I^{k+1}$ is computed through
nodal velocities $\vec{v}_I$ and accelerations $\vec{a}_I$ and evaluated
in the $k$ time step,
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{\tens{m}_I^{k+1}}\ .
\end{equation}
Where \gls{gamma-npc} is an user-defined parameter (typically adopted as
0.5), $\tens{m}_I$ represents the lumped mass matrix, and $N_{Ip}^{k}$ is
the nodal contribution of the shape function evaluated in the particle
coordinates $\vec{x}_p^k$. This way of computing the nodal predictor is numerically stable
as well as minimises the computational effort. Once nodal velocities are
obtained, the essential boundary conditions are imposed. After that, an
Eulerian phase is computed in the set of nodes in a
\acrshort{fem}-like way, where nodal forces $\vec{f}_{I}^{k+1}$ are
computed through the equilibrium equation. Next the nodal velocities
are corrected in a \textit{corrector} stage:
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\
  \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}\ .
\end{equation}
Finally, in order to update the particles, those are advected in the Lagrangian stage as
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} =
          \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ .
\end{align}
The complete pseudo-algorithm it is summarised in \ref{sec:expl-pred-corr}.

@@

(0,\infty)$. Notice that $\beta$ has units of $[L]^{-2}$,

@@

$\widehat{\gamma}=\beta h^2$, where $h$ is defined as a

@@

function is that, in practice, the value of $N_I$ decay exponentially following $\exp(-\beta \vec{r}
)$. In this sense,
a good practice is to truncate it under a tolerance  (typically around 10$^{-6}$) that would ensure a reasonable

@@

\subsection{Fracture modelling approach}
\label{sec:2.3}
Within the context of \acrshort{mpm} formulation, fracture can be
modelled by failing particles according to a suitable failure
criterion. When material points are failed, they are assumed to have
null stress tensor. To reproduce this behaviour in the present
research, the eigensoftening algorithm is introduced in the 
\acrshort{mpm} framework as an alternative approach to the decohesion
model \cite{Zhenmao_2005,Schreyer_2002}. The eigensoftening concept
was originally developed by Navas {\it et al.}
(2017)\cite{Navas_2017_ES} as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. A
comparison between both in \cite{Navas_2017_ES} shows that the
eigenerosion algorithm significantly overestimates the tensile  
stress and the strain peaks, meanwhile eigensoftening captures the forces and crack
patterns accurately. On the other hand, eigensoftening algorithm agrees
very well with experimental results in all the aspects. Furthermore,
the accuracy of this algorithm has been proven for complex fracture
patterns such as the ones obtained with fiber reinforces concrete (FRC),
\cite{Navas_2018_ES}.\\

The key idea behind the eigenerosion algorithm is the computation of the
energy-release rate attendant to the failure of material point $p$,
\begin{align}
  \label{eq:energy-release-EE}
&G_p^{k+1} = \frac{C_{\epsilon}}{m_p^{k+1}}  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q W_q^{k+1}\\
  \label{eq:mass-EE}
&m_p^{k+1} =  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q  
\end{align}
where $B_{\epsilon}(x_p^{k+1})$ is a $^n$dimensional sphere of radius $\epsilon$
centered at $x_p^{k+1}$. The particles which lie under this sphere are
known as the $\epsilon$-neighborhood of the material point (see
\cite{Pandolfi_2012}). This concept is conveniently sketched in
Figure \ref{fig:Failed-particles}.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure-Particle-failed}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Failed-particles}
\end{figure}
%%%%%%%%%%%
Other parameters are the mass of the neighborhood $m_p^{k+1}$, the
current free-energy density per unit mass  $W_q^{k+1}$ and the
normalizing constant $C_{\epsilon}$, \PNA{which also defines the $\epsilon$-neighborhood size, being $\epsilon = C_\epsilon h$.}

The failure criterion consist of considering that a material point is failed
when $G_p^{k+1}$ surpasses the critical energy release rate that
measures the material-specific energy, $G_F$. The convergence of this
approach has been analyzed by Schmidt {\it et al.}
(2009)\cite{Schmidt_2009}, who proved that $G_F$ converges to the Griffith
fracture when the discretization size tends to zero. It is necessary to
point out that, when a material point overpass the critical energy, its
contribution to the internal forces vector is set to zero, but its
contribution to the mass matrix is preserved.\\

@@

into account the gradual failure in quasi-brittle materials. The idea

@@

cohesive crack appears when the maximum tensile strength, $f_t$, is
reached. Once the crack, $w_n$, reaches the value of the
critical crack opening, $w_c$, a stress-free crack is
attained. For intermediate values a damage value between zero

@@

  B_{\epsilon}(x_p^{k+1})} m_q \tens{\sigma}_{q,I} \delta \tens{\varepsilon}_q,

@@

material point in the $\epsilon$-neighborhood. At this point, an \textit{effective} strain $\varepsilon_q$ is introduced in order to obtain the variation of
the local strain energy as $\delta W_q = \sigma_{q,1}
\delta\varepsilon_q$. Now, with the assumption that the effective

@@

Consequently, it is possible to define an equivalent critical stress at the
material point $\vec{x}_p^{k+1}$ as

@@

where $m_p$ is the total mass of the $\epsilon$-neighborhood, defined similarly than in the eigenerosion methodology (Eq.~\eqref{eq:mass-EE}).
%\begin{equation}
%  \label{eq:averaged-mass}
%  m_p = \sum_{x_q^{k+1} \in B_{\epsilon}(x_p^{k+1})} m_q.
%\end{equation}

@@

tensile strength, $f_t$. This consists of a reduction of the internal

@@

variable and the volume for each material point $p$. \PNA{$\chi_p$ is any function, taking

@@

material points), that relates the crack opening and the residual strength. To calculate it, typical shapes of cohesive laws such as linear, bilinear or exponential, may be employed.} For the case of a linear softening such the sketched one in the Figure \ref{fig:Damage-ft-wc}, $\chi_p$ is computed as,

@@

Analogous to the crack band model presented by

@@

\cite{Navas_2017_ES} introduced a band width parameter $h_{\epsilon}$ in order to relate the crack opening with the strain.
Concerning this parameter, a typical value between two and
four times the maximum size of the aggregates is adopted in the case of

@@

$\varepsilon_1(\vec{x}_p^{0})$, and the current strain, $\varepsilon_1(\vec{x}_p^{k+1})$, for a material point $p$. Also,

@@

The function of $\chi$ presented in \eqref{eq:damage-variable-chi-II}

@@

\subsection{$\epsilon$-neighborhood reconstruction : A node-linked method}
\label{sec:epsil-neighb-reconst}
\PNA{The construction of the $\epsilon$-neighborhood is a major issue of the proposed methodology.} Since this operation could be extremely demanding, optimal
numerical implementation should be employed. In opposite with the Cell-linked
method proposed by Allen \& Tildesley \cite{Allen_et_al_1989} where
the definition of a numerical tolerance is required for auxiliar
algorithms like Crossing Number \cite{Shimrat_1962} or Correct
Even-Odd \cite{Galetzka_et_al_2017}, \PNA{the proposed methodology adopted herein is linked to the shape function neighbourhood. In order to avoid the aforementioned error} prone
techniques, we endeavour to exploit the meshfree benefits of
the \acrshort{lme} approximants to introduce a node-linked method. Due
to the global definition of this interpolation technique, the classical
interpretation of the \acrshort{fem} element typically adopted in
\acrshort{mpm} simulations is not required anymore. A linked
list of tributary nodes is reconstructed with the closest node as
starting point at each time step instead. Therefore, each node is assigned to an
unique linked-list of those particles which are closest to it. This
approach allows to reconstruct easily a search list with the support
of the initial connectivity of the mesh and offers a suitable basis
for an easy implementation of other interpolation techniques with
minimum coding effort. A detailed explanation of the proposed
algorithm can be found in \ref{sec:node-linked-method-1}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cases of study and discussion}
\label{sec:3}
